Structured Synchronous Reactive Programming for Game Development \\ \Large{Case Study: On Rewriting Pingus from C++ to \CEU}

### 2017-05-24 paper.tex
...

In this work, we advocate structured synchronous reactive programming as an
expressive and productive alternative for game logic development.
We present a case study of rewriting Pingus from C++ to \CEU with the
contributions as follows:
%
\begin{itemize}
\item Applying idiomatic code in \CEU as alternative solutions for six selected
      behaviors in the game logic.
\item Presenting an in-depth qualitative analysis of the proposed solutions in
      comparison to the original implementations in C++.
\item Identifying four recurrent control-flow patterns that likely apply to
      other games:
        \emph{Finite State Machines},
        \emph{Continuation Passing},
        \emph{Dispatching Hierarchies}, and
        \emph{Lifespan Hierarchies}.
        %\emph{Signaling Mechanisms}.
\end{itemize}
%
A control-flow pattern is a recurring technique to describe dependencies and
explicit orders between statements (or groups of statements) in a program.
For instance, consider how a key press stimulus propagates through the game
entities and also what happens with them if the stimulus causes the end of the
game.
%
\CEU supports primitives that help describing complex control-flow
relationships in the game logic more concisely.
%
The rewriting process consisted of identifying sets of callbacks implementing
control flow in the game and translating them to \CEU using appropriate
structured constructs.
%
As an example, a double mouse click is characterized by a first click, followed
by a maximum amount of time, followed by a second click.
This behavior depends on different events (clicks and timers) which have to
occur in a particular order.
In C++, the implementation involves callbacks crossing reactions to successive
events which manipulate state variables explicitly.
%
We only present six cases distributed in the four patterns due to space
constraints, but we actually documented in detail a total eleven cases in six
patterns (we omit the sixth pattern \emph{Signaling Mechanisms} entirely).
%
This work focuses on a qualitative analysis for the programming techniques
that we applied during the rewriting process.
Not all techniques result in reduction in LoCs (especially considering the
verbose syntax of \CEU), but have other effects such as eliminating shared
variables and dependencies between classes.

@@

\subsubsection{Case Study: The \emph{Bomber Action} Animation Sequence}

@@

\subsubsection{Case Study: Managing the Pingus Lifecycle}

@@

TODO: case 2

>>>>

In this work, we advocate structured synchronous reactive programming as an
expressive and productive alternative for game logic development.
We present a case study of rewriting Pingus from C++ to \CEU with the
contributions as follows:
%
\begin{itemize}
\item Applying idiomatic code in \CEU as alternative solutions for six selected
      behaviors in the game logic.
\item Presenting an in-depth qualitative analysis of the proposed solutions in
      comparison to the original implementations in C++.
\item Identifying four recurrent control-flow patterns that likely apply to
      other games:
        \emph{Finite State Machines},
        \emph{Continuation Passing},
        \emph{Dispatching Hierarchies}, and
        \emph{Lifespan Hierarchies}.
        %\emph{Signaling Mechanisms}.
\end{itemize}
%
A control-flow pattern is a recurring technique to describe dependencies and
explicit orders between statements (or groups of statements) in a program.
For instance, consider how a key press stimulus propagates through the game
entities and also what happens with them if the stimulus causes the end of the
game.
%
\CEU supports primitives that help describing complex control-flow
relationships in the game logic more concisely.
%
The rewriting process consisted of identifying sets of callbacks implementing
control flow in the game and translating them to \CEU using appropriate
structured constructs.
%
As an example, a double mouse click is characterized by a first click, followed
by a maximum amount of time, followed by a second click.
This behavior depends on different events (clicks and timers) which have to
occur in a particular order.
In C++, the implementation involves callbacks crossing reactions to successive
events which manipulate state variables explicitly.
%
We only present six cases distributed in the four patterns due to space
constraints, but we actually documented in detail a total of eleven cases in six
patterns (we omit the sixth pattern \emph{Signaling Mechanisms} entirely).
%
This work focuses on a qualitative analysis for the programming techniques
that we applied during the rewriting process.
Not all techniques result in reduction in LoCs (especially considering the
verbose syntax of \CEU), but have other effects such as eliminating shared
variables and dependencies between classes.

@@

\subsubsection{Case Study: The \emph{Bomber Action} Animation Sequence}
\label{sec.pats.fsms.2}

@@

\subsubsection{Case Study: Game UI Widgets}

\begin{figure}[t]
\centering
\includegraphics[width=\columnwidth]{game-session-arrows}
\caption{UI children with static lifespan.
\label{fig.ui}
}
\end{figure}

\begin{figure*}[t]
\begin{minipage}[t]{0.54\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
GameSession::GameSession(<...>) :
    <...>
{
    <...>      // these widgets are always active...
    btpanel  = new ButtonPanel(<...>);
    pcounter = new PingusCounter(<...>);
    smallmap = new SmallMap(<...>);
    <...>
    uimgr->add(btpanel);  // ...but are added
    uimgr->add(pcounter); //  dynamically to the
    uimgr->add(smallmap); //  dispatching hierarchy
    <...>
}
\end{lstlisting}
\centering\small{\ax Implementation in C++}
\end{minipage}
%
\begin{minipage}[t]{0.46\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
code/await Game (void) do
    <...> // other coexisting functionality
    spawn ButtonPanel(<...>);
    spawn PingusCounter(<...>);
    spawn SmallMap(<...>);
    <...> // other coexisting functionality
end
\end{lstlisting}
\centering\small{\bx Implementation in \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Managing the UI widgets lifecycle.
\label{lst.pingus}
}
\end{figure*}

Figure~\ref{fig.ui} shows the game UI widgets holding the buttons, pingus
counter, and small map which coexist with the game screen during its whole
lifespan.
%
Figure~\ref{lst.ui} compares the implementations in C++ and \CEU.

In C++, the widgets are created in the constructor of the class
\code{GameSession} (ln. 5--7), added to a UI container (ln. 9--11), and are
never removed since they must always be visible.
Arguably, to better express the intent of making them coexist with the game
screen, the widgets could be declared as top-level automatic (non-dynamic)
members.
However, the class uses a container to automate \code{draw} and \code{update}
dispatching to the widgets, as discussed in Section~\ref{sec.pats.dispatching}.
In turn, the container method \code{add} expects dynamically allocated children
only because they are automatically deallocated inside the container
destructor.
%
The dynamic nature of containers in C++ demand extra caution from programmers:
%
\begin{itemize}
\item When containers are part of a dispatching chain, it gets even harder to
      know which objects are dispatched:
      one has to ``simulate'' the program execution and track calls to
      \code{add} and \code{remove}.
\item For objects with dynamic lifespan, calls to \code{add} must always have
      matching calls to \code{remove}:
      missing calls to \code{remove} lead to memory and CPU leaks (to be
      discussed as the \emph{lapsed listener} problem in
      Section~\ref{sec.pats.lifespan.2}).
\end{itemize}

In Céu, entities that coexist just have to be created in the same lexical
block.
%
Since abstractions can react independently, they do not require a dispatching
container.
%
Lexical lifespan never requires containers, allocation and deallocation, or
explicit references.
In addition, all required memory is known at compile time, similarly to
stack-allocated local variables.
%
The \emph{Bomber action} of Section~\ref{sec.pats.fsms.2} also relies on
lexical scope to delimit the lifespan of the explosion sprite to a single
frame.

\subsubsection{Case Study: Managing the Pingus Lifecycle}
\label{sec.pats.lifespan.2}

@@
